Nathan Greenslit Data Analysis Project

### 2023-04-18 code/analysis_code/stats/3_correlation.qmd
edits
Made spearman and pearson Cor plots

[**THIS CODE IS STILL IN THE WORKS**]{.ul}

@@

### Log10 Transform non-normally distributed data and Remove Non-Parametric Variables

<!--# If we are using a nonparametric correlation (i.e. Spearman) we do not log-transform data -->

@@

```{r}
#Blind Oso
#bo<-
  #bo %>%
 # mutate(amm = log10(amm), #Log-transforms data
        # din_dip = log10(din_dip),
        # doc = log10(doc),
        # dust = log10(dust),
         #chl = log10(chl)) %>%
 # select(!c(secchi, tn, tdn, don)) #Removes non-parametric data

#Canals
#c2<- 
  #c2 %>%
  #mutate(orthop = log10(orthop),
         #dust = log10(dust)) %>%
  #select(!c(secchi, nn, sil, doc,toc, tn))

#Gulf
#rd<- 
 # rd %>%
 # mutate(doc = log10(doc),
      #   toc = log10(toc),
        # dust = log10(dust)) %>%
  #select(!c(sal, do_mgl, secchi, nn, tn))
```

@@

```{r, include = FALSE}
b<- cor(bo, method = "spearman") #Default method = pearson
round(b,2) #Round correlation coeff to "x" digits

c<- cor(c2, method = "spearman")
round(c,2)

r<- cor(rd, method = "spearman")
round(r,2)
```

@@

```{r}
bclean<- rcorr(as.matrix(bo))
flattenCorrMatrix(bclean$r, bclean$P)

cclean<- rcorr(as.matrix(c2))
flattenCorrMatrix(cclean$r, cclean$P)

rclean<- rcorr(as.matrix(rd))
flattenCorrMatrix(rclean$r, rclean$P)
```

@@

<!--# This is using Spearman's with raw data -->

```{r}
corrplot(b, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Blind Oso",  mar=c(0,0,1,0)) #Adds and lowers title

corrplot(c, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Canals", mar=c(0,0,1,0))

corrplot(r, type = "upper", 
         tl.col = "black", tl.srt = 45, mar=c(0,0,1,0))
         title(main = "Gulf" )
```

>>>>

@@

#### Make Datasets for normal and non-normal variables

Variables that are normally distributed across all three sites will be in one df. All others will be in another.

```{r}
b_norm<- 
  bo %>%
  select(!c(sal, secchi, nn, tdn, don, doc, toc, tn))
b_para<- 
  bo %>%
  select(sal, secchi, nn, tdn, don, doc, toc, tn)

c_norm<- 
  c2 %>%
  select(!c(sal, secchi, nn, tdn, don, doc, toc, tn))
c_para<- 
  c2 %>%
  select(sal, secchi, nn, tdn, don, doc, toc, tn)

r_norm<- 
  rd %>%
  select(!c(sal, secchi, nn, tdn, don, doc, toc, tn))
r_para<- 
  rd %>%
  select(sal, secchi, nn, tdn, don, doc, toc, tn)
```

### Log10 Transform non-normally distributed data for Norm df

@@

```{r}
#Blind Oso
b_norm<-
  b_norm %>%
  mutate(amm = log10(amm), #Log-transforms data
         din_dip = log10(din_dip),
         dust = log10(dust),
         chl = log10(chl))

#Canals
c_norm<- 
  c_norm %>%
  mutate(orthop = log10(orthop),
         dust = log10(dust))

#Gulf
r_norm<- 
  r_norm %>%
  mutate(dust = log10(dust))
```

@@

```{r, include = FALSE}
#Normal Data = Pearson's
b<- cor(b_norm, method = "pearson") #Default method = pearson
round(b,2) #Round correlation coeff to "x" digits

c<- cor(c_norm, method = "pearson")
round(c,2)

r<- cor(r_norm, method = "pearson")
round(r,2)

#Non-parametric Data  = Spearman's
b1<- cor(b_para, method = "spearman") 
round(b,2) #Round correlation coeff to "x" digits

c1<- cor(c_para, method = "spearman")
round(c,2)

r1<- cor(r_para, method = "spearman")
round(r,2)
```

@@

```{r}
bclean<- rcorr(as.matrix(b_norm))
flattenCorrMatrix(bclean$r, bclean$P)

cclean<- rcorr(as.matrix(c_norm))
flattenCorrMatrix(cclean$r, cclean$P)

rclean<- rcorr(as.matrix(r_norm))
flattenCorrMatrix(rclean$r, rclean$P)
```

@@

##### Normal

```{r}
corrplot(b, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Blind Oso_Pearson",  mar=c(0,0,1,0)) #Adds and lowers title

corrplot(c, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Canals_Pearson", mar=c(0,0,1,0))

corrplot(r, type = "upper", 
         tl.col = "black", tl.srt = 45, mar=c(0,0,1,0))
         title(main = "Gulf_Pearson" )
```

##### Non-Para

```{r}
corrplot(b1, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Blind Oso_Spearman",  mar=c(0,0,1,0)) #Adds and lowers title

corrplot(c1, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Canals_Spearman", mar=c(0,0,1,0))

corrplot(r1, type = "upper", 
         tl.col = "black", tl.srt = 45, mar=c(0,0,1,0))
         title(main = "Gulf_Spearman" )
```

### 2023-04-18 products/manuscript/Manuscript.qmd
Made spearman and pearson Cor plots

1.  I would do a lot of visual inspections to see how the different models fit and what they predict. You can try a linear model and a more complex ML one and compare to see if maybe something beyond a linear relationship improves results.
2.  Given that you don't have a lot of data, I'm not sure a PCA makes sense. I think in-depth analysis of the relation between the variables and the outcome and then in addition to the linear models trying a few more complex ones will probably be the better approach.
3.  The figures in the manuscript look good, but some are too small to read. Needs breaking up and/or increasing font size

@@

Dust concentration (AOT) was collected at time points 0 hour, 6 hour, 12 hour, and 18 hour. In this script, I made different data sets based on time points as well as a data set containing the summed dust concentration (of all time points) per day. For downstream statistical analysis, it will be helpful to have these different data sets to compare to time of sample collection.

@@

@@

The top portion of the figure depicts total enumerated Vibrio as copies per mL, with color by site. There is a noticeable shift in copies in the Gulf site, a less noticeable but still present shift in Blind Oso Bay, and practically no changes in the Canal site. The bottom portion of the figure depicts the dust concentration (AOD) across the daily time series. This figure depicts the summed dust concentration 24hr prior to sample collection (time points 13hr--> 7hr the next day) across the daily time series. During the time series, two small periods of higher dust AOD occurred on the 9th and 13th respectively. A much larger spike in AOD was observed on the 16th. Initial shift in copies per mL at the Gulf and Blind Oso occur 24-48hr following initial dust introduction on the 13th, and experience another shift around 24-48hr of higher dust input on the 16th.

@@

### Correlation Matirices

@@

This study aimed to examine microbial response to Saharan dust deposition in high nutrient coastal waters. Specifically focusing on bacteria belong to the genus *Vibrio*, we found that growth response following dust input varied by site. This could be due to differing site dynamics such as salinity, temperature, and nutrient levels.

@@

For this analysis, we examined *Vibrio* growth dynamics in response to Saharan Dust depsoition in three sites in Corpus Christi, TX that had differing baseline nutrient levels. The Gulf, with relatively constant salinities and low nutrient levels, Blind Oso Bay with high nutrients and fluctuating salinities, and a residential canal system with consistent temperature and salinities and nutrient levels. Samples were collected before, during, and after a dust event and *Vibrio* were enumerated using qPCR. Data on nutrient levels and other environmental parameters (e.g. salinity and temperature) were also collected during this period. The three data sets (Vibrio Enumeration, Environmental parameters, and Dust Concentration) were then processed and cleaned for analysis.

@@

#### Dust

The daily time series captured two dust events. One light event starting on the 13th of July, where AOD reached 0.2, and a much heavier dust event from the 16th to the 18th of July with AOD = 55-57. Dust AOD returned to lower concentrations following the time series.

@@

Peaks in *Vibrio* growth were observed at the Gulf 1-2 days following dust deposition. Blind Oso exhibited a similar response, but was slightly dampened in comparison. This could be due to the site-specific nutrient dynamics. While the Gulf exhibited a sharper increase during the second, heavy dust event, Blind Oso exhibited a smaller increase as compared to the light dust event. \
\
If the Gulf has presumably lower nutrient levels, the constituents from the influx of dust may be used up rapidly, so we still see a sharp secondary growth response at the Gulf during the heavy dust event. Blind Oso Bay has presumably higher nutrient levels, so while we see an initial spike in growth, the secondary spike is not as dramatic. So in this case, it may be that the dust has reached a maximum to how much it can support growth, possibly due to the dust constituents not being as essential for growth at this inland site.

#### Statistical Analysis

@@

>>>>

1.  I would do a lot of visual inspections to see how the different models fit and what they predict. You can try a linear model and a more complex ML one and compare to see if maybe something beyond a linear relationship improves results.

2.  Given that you don't have a lot of data, I'm not sure a PCA makes sense. I think in-depth analysis of the relation between the variables and the outcome and then in addition to the linear models trying a few more complex ones will probably be the better approach.

    <!--# Try a model that is copies_mL ~. -->

3.  The figures in the manuscript look good, but some are too small to read. Needs breaking up and/or increasing font size

@@

Dust concentration (AOT) was collected at time points 0 hour, 6 hour, 12 hour, and 18 hour. In this script, I converted timpoints from UTC to CST/CDT and created different dust permutations (individual time points, sums by day, and average by day). Because samples were collected between 0700 and 1100, a 24h composite permutation was made. This permutation is what was used for all downstream analyses.

@@

This figure depicts the summed dust concentration 24hr prior to sample collection (time points 13hr--> 7hr the next day) across the daily time series. The daily time series captured two dust events. One light event starting on the 13th of July, where AOD reached 0.2, and a much heavier dust event from the 16th to the 18th of July with AOD = 55-57. Dust AOD returned to lower concentrations following the time series. The top portion of the figure depicts total enumerated Vibrio as copies per mL, with color by site. There is a noticeable shift in *Vibrio* growth at the Gulf site, a less noticeable but still present shift in Blind Oso Bay, and practically no changes in the Canal site. An initial shift in copies per mL at the Gulf and Blind Oso occur 24-48hr following initial dust introduction on the 13th, and experience another shift around 24-48hr of higher dust input on the 16th

@@

@@

### Correlation Matrices

@@

This study aimed to examine microbial response to Saharan dust deposition in high nutrient coastal waters. Specifically focusing on bacteria belong to the genus *Vibrio*, we found that growth response following dust input varied by site.

@@

For this analysis, we examined *Vibrio* growth dynamics in response to Saharan Dust deposition in three sites in Corpus Christi, TX that had differing baseline nutrient levels. The Gulf, with relatively constant salinities and low nutrient levels, Blind Oso Bay with high nutrients and fluctuating salinities, and a residential canal system with consistent temperature and salinities and nutrient levels. Samples were collected before, during, and after a dust event and *Vibrio* were enumerated using qPCR. Data on nutrient levels and other environmental parameters (e.g. salinity and temperature) were also collected during this period. The three data sets (Vibrio Enumeration, Environmental parameters, and Dust Concentration) were then processed and cleaned for analysis.

@@

@@

Overall, *Vibrio* growth was observed following dust events at both the Gulf and Blind Oso. Peaks in growth were observed 24-48 hours following dust deposition. Linear regression analysis confirmed a correlation between growth and dust input (R2 = 0.49, 0.56, 0.44 for Blind Oso, Canals, and Gulf respectively). However, the environmental dynamics of each site most likely contributed the degree of response per site. Multivariate models were tested using specific environmental parameters: dust, DOM (DON, DOC), temperature, salinity, orthophosphate, and nitrate+nitrite. Ultimately, a model that included water chemistry, dust, nutrient, and dissolved organic matter parameters provided the best performance (R2 = 0.73, 0.7, and 0.76 for Blind Oso, Canals, and Gulf respectively), suggesting that while dust may contribute to a growth response in *Vibrio* following dust input, the site specific dynamics aid in dictating the degree of response observed.

The Canals may have a heavily dampened growth response due to a high and constant salinity of 40-42. In general, optimal growth occurs at salinities of \~25-35, and less tolerant species will drop as salinities climb <!--# CITE --> .

Additionally, the Gulf exhibited a sharper increase during the second, heavier dust event, while Blind Oso exhibited a smaller increase during the heavy dust event. This difference in response may be attributed to the nutrient dynamics of the sites. In marine waters, nutrients are typically limiting due to tight microbial recycling <!--# CITE -->. Therefore the introduction on nutrients at such a site may result in rapid uptake and utilization. At Blind Oso, these dust constituents may not be as essential for growth, and the dust may have reached a maximum in how much it can support growth.

@@

At Blind Oso for example, the salinity is observed to drop from 42-20 within an eight day period that coincided with dust input. It is difficult to parse these two variables out as either or both could have contributed to growth at this site.
